Legion: Expressing Locality and Independence with Logical Regions

### 2012-07-03 doc/popl2013/langdef.tex
checkpoint - done with first draft text through soundness proof

@@

\begin{figure*}
\centering{
\framebox{$\typeenv{bv}{bool}$}
\framebox{$\typeenv{iv}{int}$}
\finfrule
{\begin{array}{l}
\typeenvx e_1 : T_1 \\
\typeenvx e_2 : T_2
\end{array}}
{\typeenvx \langle e_1, e_2 \rangle : \langle T_1, T_2 \rangle}
\finfrule{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.1}\ : T_1}
\finfrule{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.2}\ : T_2}
\finfrule{\Gamma(id) = T}{\typeenvx id : T}
\framebox{$\typeenvx \text{null }T@r : T@r$}
\framebox{$\typeenvx \text{new }T@r : T@r$}
\finfrule{\typeenvx e : T@(\oton{r})}{\typeenvx \text{isnull}(e) : bool}
\finfrule
{\begin{array}{l}
\typeenvx e : T@(\otok{k}{r'}) \\
\forall i. \exists j, r'_i \leq r_j \in \Omega^* \\
\end{array}}
{\typeenvx upregion(e,\oton{r}) : T@(\oton{r})}
\finfrule
{\begin{array}{l}
\typeenvx e : T@(\otok{k}{r'}) \\
\forall j. \exists i, r_j \leq r'_i \in \Omega^* \\
\end{array}}
{\typeenvx downregion(e,\oton{r}) : T@(\oton{r})}
\finfrule
{\begin{array}{l}
\typeenvx e_1 : T@(\oton{r}) \\
\forall i, reads(r_i) \in \Phi^*\end{array}}
{\typeenvx \text{read}(e_1) : T}
\finfrule
{\begin{array}{l}
\typeenvx e_1 : T@(\oton{r}) \\
\typeenvx e_2 : T \\
\forall i, writes(r_i) \in \Phi^*
\end{array}}
{\typeenvx \text{write}(e_1, e_2) : bool}
\finfrule
{\begin{array}{l}
\Gamma(id) = (T_1, T_2), \emptyset, \emptyset \rightarrow T_1 \\
\typeenvx e_1 : T_1@(\oton{r}) \\
\typeenvx e_2 : T_2 \\
\forall i, reduces(id, r_i) \in \Phi^*
\end{array}}
{\typeenvx \text{reduce}(id, e_1, e_2) : bool}
\framebox{$\typeenvx \text{newcolor }r : \text{coloring}(r)$}
\finfrule{\begin{array}{l}
\typeenvx e_1 : \text{coloring}(r) \\
\typeenvx e_2 : T@r \\
\typeenvx e_3 : int
\end{array}}
{\typeenvx \text{color}(e_1, e_2, e_3) : \text{coloring}(r)}
\finfrule{\begin{array}{l}\typeenvx e_1 : int \\ \typeenvx e_2 : int\end{array}}{\typeenvx e_1 + e_2 : int}
\finfrule{\begin{array}{l}\typeenvx e_1 : int \\ \typeenvx e_2 : int\end{array}}{\typeenvx e_1 < e_2 : bool}
\finfrule{\begin{array}{l}
\typeenvx e_1 : T_1 \\
\Gamma' = \Gamma[id/T_1] \\
\typeenvx[G=\Gamma'] e_2 : T_2
\end{array}}
{\typeenvx : \text{let}\ id : T_1 \text{in}\ e_2 : T_2}
\finfrule{\begin{array}{l}\typeenvx e_1 : bool \\ \typeenvx e_2 : T \\ \typeenvx e_3 : T\end{array}}{\typeenvx \text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3 : T}
\finfrule{
\begin{array}{l}
\Gamma(id) = \forall \otok{k}{r'}.(\oton{T}),\Phi_i, Q \rightarrow T_r \\
\typeenvx e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\
\Phi_i[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*
\end{array}}
{\typeenvx id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]}
\finfrule{
\begin{array}{l}
\typeenvx e_1 : \text{coloring}(r_p) \\
\Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{i \not= j} r_i * r_j \\
\typeenvx[O=\Omega'] e_2 : T \\
r_i \not\in \textit{regions\_of}(\Gamma, T)
\end{array}}
{\typeenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 : T}
\finfrule{
\begin{array}{l}
T_1 = \exists \otok{k}{r'}. T_2\text{ where }\Omega_1 \\
\Omega \models \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
\typeenvx e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]
\end{array}}
{\typeenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1}
\finfrule{
\begin{array}{l}
T_1 = \exists \otok{k}{r'}. T_2\text{ where }\Omega_1 \\
\typeenvx e_1 : T_1 \\
\Gamma' = \Gamma[T_2[r_1/r'_1,\ldots,r_k/r'_k] / id] \\
\Omega' = \Omega \wedge \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
\typeenvx[G=\Gamma',O=\Omega'] e_2 : T_3 \\
r_i \not\in \textit{regions\_of}(\Gamma, T_1, T_3)
\end{array}}
{\typeenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 : T_3}
}
\caption{Legion Core Type System}
\label{fig:types}
\end{figure*}

@@

\opsenvx[L=L',S=S'] e_2 \mapsto v_2, E_2

@@

\opsenvx[L=L',S=S'] e_3 \mapsto v_3, E_3

@@

An important property of the legion type system is that any expression that is well-typed is
guaranteed to access the heap only in ways permitted by the privileges under which it was typed.
A judgement $M \vdash E : \Phi$
If
\begin{enumerate}
\item $\typeenvx e : T$
\item $\opsenvx e \mapsto v, E$
\end{enumerate}
we would like to prove that all of the dynamic effects described by $E$ are
included in the static privileges $\Phi$.  A judgement $E : \Phi$ means that
for all effects $\epsilon$ in $E$,
\begin{enumerate}
\item if $\epsilon = read(l, v)$, then for some $r$, $reads(r) \in \Phi \wedge l \in M(r)$
\item if $\epsilon = write(l, v)$, then for some $r$, $writes(r) \in \Phi \wedge l \in M(r)$
\item if $\epsilon = reduce(id, l, v)$, then for some $r$, $reduces(id,r) \in \Phi \wedge l \in M(r)$
\end{enumerate}

\subsection{Strategy}

The mapping $M$ is a function from logical region names $r$ to physical regions $\rho$.  It will
be convenient to apply $M$ to types and even functions that operate on types by defining
$\hat x = M \llbracket x \rrbracket$ to be $x$ with all logical region names $r_i$ replaced by their
corresponding physical region $\rho_i = M(r_i)$.   \\

Soundness is proven using induction over the structure of an expression.  A number of properties
are assumed and/or proven at each step: 
\begin{enumerate}
\item $M \sim \Omega \Leftrightarrow (\forall r_i, r_j. r_i \leq r_j \in \Omega \rightarrow M(r_i) \subseteq M(r_j)) \wedge (\forall r_i, r_j. r_i * r_j \in \Omega \rightarrow M(r_i) \cap M(r_j) = \emptyset)$, that M is an $\Omega$-satisfying mapping of logical regions to physical regions
\item $\hat T = M \llbracket T \rrbracket$, a {\em physical type}, constructed by replacing all of the free region names $r$ in T with the physical regions $\rho = M(r)$
\item $v \sim_H \hat T$, value consistency with respect to a heap typing.  Value consistency is
itself defined inductively over the structure of a type:
\begin{enumerate}
\item $bv \sim_H bool$
\item $iv \sim_H int$
\item $null \sim_H T@\rho$
\item $l \sim_H T@\rho \Leftrightarrow (l \in \rho) \wedge (H(l) \sim_H T)$
\item $\langle v_1, v_2 \rangle \sim_H \langle \hat T_1, \hat T_2 \rangle \Leftrightarrow (v_1 \sim_H \hat T_1) \wedge (v_2 \sim_H \hat T_2)$
\item $\langle \langle \oton{\rho}, v \rangle \rangle \sim_H \text{rr}[\oton{r}] \hat T\text{ where }\Omega \Leftrightarrow (v \sim_H \hat T[\rho_1/r_1, \ldots \rho_n/r_n]) \wedge (\{ (r_i, \rho_i) \} \sim \Omega)$
\item $K \sim_H coloring(\rho) \Leftrightarrow \forall l_1, v_1. (l_1, v_1) \in K \rightarrow (l_1 \in \rho \wedge \forall l_2, v_2. (l_2, v_2) \in K \rightarrow (l_1 \not= l_2) \vee (v_1 = v_2))$
\end{enumerate}
\item $L \sim_H M \llbracket \Gamma \rrbracket \Leftrightarrow \forall (id,v) \in L. v \sim_H M \llbracket \Gamma \rrbracket (id) $, local stack consistency with respect to a heap typing
\item $S \sim H \Leftrightarrow \forall (l,v) \in S. v \sim_H H(l)$, store consistency with respect to a given heap typing
\item $E \sim H \Leftrightarrow (\forall l, v. write(l,v) \in E \rightarrow v \sim_H H(l)) \wedge (\forall id, l, v. reduce_{id}(l, v) \in E \rightarrow (\hat \Gamma (id) = (\hat T_1, \hat T_2), \emptyset \rightarrow \hat T_1) \wedge H(l) = \hat T_1 \wedge v \sim_H \hat T_2)$, effect consistnecy with respect to a given heap typing
\end{enumerate}

Armed with all of these, we look at the following property:
\begin{enumerate}
\item $E \sim_H \Phi$, containment of effects within privileges
\end{enumerate}

and prove

\finfrule
{\begin{array}{l}
\typeenvx e : T \\
\opsenvx e \mapsto v, E \\
M \sim \Omega \hspace{1cm}
L \sim_H \Gamma \hspace{1cm}
S \sim H
\end{array}}
{
v \sim_H M \llbracket T \rrbracket \hspace{1cm}
E \sim H \hspace{1cm}
E : \Phi
}

A few lemmas will be useful:

\finfrule
{\begin{array}{l}
S \sim H \hspace{1cm}
E \sim H
\end{array}}
{ \text{apply}(S, E, C) \sim H }

\finfrule
{\begin{array}{l}
E_1 \sim H \hspace{1cm}
E_2 \sim H
\end{array}}
{ E_1 \dplus E_2 \sim H } \\

\finfrule
{\begin{array}{l}
E_1 \sim H \hspace{1cm}
E_2 \sim H \\
valid\_interleave(S, C, E', E_1, E_2)
\end{array}}
{ E' \sim H } \\

\finfrule
{\begin{array}{l}
E_1 : \Phi \hspace{1cm}
E_2 : \Phi
\end{array}}
{ E_1 \dplus E_2 : \Phi } \\

\framebox{$M \sim \Omega \Leftrightarrow M \sim \Omega^*$} \\

\framebox{$E : \Phi \Leftrightarrow E : \Phi^*$}

\begin{table*}
\centering
{\small
\begin{tabular}{l|p{5cm}|p{5cm}|p{5cm}}
{\bf Rule} & 
\multicolumn{1}{|c|}{$\mathbf{v \sim_H M \llbracket T \rrbracket}$} &
\multicolumn{1}{|c|}{$\mathbf{E \sim H}$} &
\multicolumn{1}{|c|}{$\mathbf{E : \Phi}$} \\
\hline
BoolConst & & & \\
IntConst & & & \\
TupleBuild & & & \\
TupleField & & & \\
LocalVar & & & \\
NullPtr & & & \\
NewPtr & & & \\
IsNull & & & \\
UpRegion & & & \\
DownRegion & & & \\
Read & & & \\
Write & & & \\
Reduce & & & \\
Newcolor & & & \\
Color & & & \\
IntMath & & & \\
IntCmp & & & \\
Let & & & \\
IfElse & & & \\
TaskCall & & & \\
Partition & & & \\
Pack & & & \\
Unpack & & &
\end{tabular}}
\caption{Proof pieces}
\end{table*}

>>>>

\newtheorem{thm}{Theorem}
\newtheorem{lem}{Lemma}

@@

\begin{figure*}
\centering{
\framebox{$\typeenv{bv}{bool}$}
\framebox{$\typeenv{iv}{int}$}
\finfrule
{\begin{array}{l}
\typeenvx e_1 : T_1 \\
\typeenvx e_2 : T_2
\end{array}}
{\typeenvx \langle e_1, e_2 \rangle : \langle T_1, T_2 \rangle}
\finfrule{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.1}\ : T_1}
\finfrule{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.2}\ : T_2}
\finfrule{\Gamma(id) = T}{\typeenvx id : T}
\framebox{$\typeenvx \text{null }T@r : T@r$}
\framebox{$\typeenvx \text{new }T@r : T@r$}
\finfrule{\typeenvx e : T@(\oton{r})}{\typeenvx \text{isnull}(e) : bool}
\finfrule
{\begin{array}{l}
\typeenvx e : T@(\otok{k}{r'}) \\
\forall i. \exists j, r'_i \leq r_j \in \Omega^* \\
\end{array}}
{\typeenvx upregion(e,\oton{r}) : T@(\oton{r})}
\finfrule
{\begin{array}{l}
\typeenvx e : T@(\otok{k}{r'}) \\
\forall j. \exists i, r_j \leq r'_i \in \Omega^* \\
\end{array}}
{\typeenvx downregion(e,\oton{r}) : T@(\oton{r})}
\finfrule
{\begin{array}{l}
\typeenvx e_1 : T@(\oton{r}) \\
\forall i, reads(r_i) \in \Phi^*\end{array}}
{\typeenvx \text{read}(e_1) : T}
\finfrule
{\begin{array}{l}
\typeenvx e_1 : T@(\oton{r}) \\
\typeenvx e_2 : T \\
\forall i, writes(r_i) \in \Phi^*
\end{array}}
{\typeenvx \text{write}(e_1, e_2) : bool}
\finfrule
{\begin{array}{l}
\Gamma(id) = (T_1, T_2), \emptyset, \emptyset \rightarrow T_1 \\
\typeenvx e_1 : T_1@(\oton{r}) \\
\typeenvx e_2 : T_2 \\
\forall i, reduces(id, r_i) \in \Phi^*
\end{array}}
{\typeenvx \text{reduce}(id, e_1, e_2) : bool}
\framebox{$\typeenvx \text{newcolor }r : \text{coloring}(r)$}
\finfrule{\begin{array}{l}
\typeenvx e_1 : \text{coloring}(r) \\
\typeenvx e_2 : T@r \\
\typeenvx e_3 : int
\end{array}}
{\typeenvx \text{color}(e_1, e_2, e_3) : \text{coloring}(r)}
\finfrule{\begin{array}{l}\typeenvx e_1 : int \\ \typeenvx e_2 : int\end{array}}{\typeenvx e_1 + e_2 : int}
\finfrule{\begin{array}{l}\typeenvx e_1 : int \\ \typeenvx e_2 : int\end{array}}{\typeenvx e_1 < e_2 : bool}
\finfrule{\begin{array}{l}
\typeenvx e_1 : T_1 \\
\Gamma' = \Gamma[id/T_1] \\
\typeenvx[G=\Gamma'] e_2 : T_2
\end{array}}
{\typeenvx : \text{let}\ id : T_1 \text{in}\ e_2 : T_2}
\finfrule{\begin{array}{l}\typeenvx e_1 : bool \\ \typeenvx e_2 : T \\ \typeenvx e_3 : T\end{array}}{\typeenvx \text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3 : T}
\finfrule{
\begin{array}{l}
\Gamma(id) = \forall \otok{k}{r'}.(\oton{T}),\Phi_i, Q \rightarrow T_r \\
\typeenvx e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\
\Phi_i[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*
\end{array}}
{\typeenvx id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]}
\finfrule{
\begin{array}{l}
\typeenvx e_1 : \text{coloring}(r_p) \\
\Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{i \not= j} r_i * r_j \\
\typeenvx[O=\Omega'] e_2 : T \\
r_i \not\in \textit{regions\_of}(\Gamma, T)
\end{array}}
{\typeenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 : T}
\finfrule{
\begin{array}{l}
T_1 = \exists \otok{k}{r'}. T_2\text{ where }\Omega_1 \\
\Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^* \\
\typeenvx e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]
\end{array}}
{\typeenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1}
\finfrule{
\begin{array}{l}
T_1 = \exists \otok{k}{r'}. T_2\text{ where }\Omega_1 \\
\typeenvx e_1 : T_1 \\
\Gamma' = \Gamma[T_2[r_1/r'_1,\ldots,r_k/r'_k] / id] \\
\Omega' = \Omega \wedge \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
\typeenvx[G=\Gamma',O=\Omega'] e_2 : T_3 \\
r_i \not\in \textit{regions\_of}(\Gamma, T_1, T_3)
\end{array}}
{\typeenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 : T_3}
}
\caption{Legion Core Type System}
\label{fig:types}
\end{figure*}

@@

\opsenvx[S=S'] e_2 \mapsto v_2, E_2

@@

\opsenvx[S=S'] e_3 \mapsto v_3, E_3

@@

An important property of the legion type system is that any expression that is well-typed is
guaranteed to access the heap only in ways permitted by the privileges under which it was typed.
A judgement $M \vdash E : \Phi$ holds if all the memory operations in $E$ are of the types and
to locations covered by the privileges in $\Phi$:

\begin{center}
\begin{tabular}{l@{ }l@{ }l}
$M \vdash E : \Phi$ & $\Leftrightarrow$ & $\forall \epsilon\text { in } E.$ \\
& & $(\epsilon = read(l, c, v) \rightarrow \exists r. l \in M(r) \wedge reads(r) \in \Phi)\ \wedge$ \\
& & $(\epsilon = write(l, c, v) \rightarrow \exists r. l \in M(r) \wedge writes(r) \in \Phi)\ \wedge$ \\
& & $(\epsilon = reduce_{id}(l, c, v) \rightarrow \exists r. l \in M(r) \wedge reduces(id, r) \in \Phi)$
\end{tabular}
\end{center}

The proof that $M \vdash E : \Phi$ holds for all well-typed Legion expressions is achieved by
induction over the structure of the expression.  Soundness is contingent on the following
consistent properties, which must be shown to hold for each subexpression:
\begin{itemize}
\item $M \sim \Omega$, a region mapping must satisfy the region constraints under which the expression was typed
\item $L \sim_H \Gamma$, the local values must have types consistent with the environment under which the expression was typed
\item $S \sim H$, the initial state of the store must contain values whose types are consisten with the heap typing
\end{itemize}

The proof requires two additional properties be proved for each subexpression:
\begin{itemize}
\item $v \sim_H M \llbracket T \rrbracket$, a subexpresion must evaluate to a value of the right type
\item $E \sim H$, the values written to the store must match the heap types of the location to which they're stored
\end{itemize}
Figure~\ref{fig:constprop} gives the precise definition of these properties.

\begin{figure*}
\centering{\footnotesize
$
\begin{array}{l}
\begin{array}{ll}
\begin{array}{lll}
M \sim \Omega & \Leftrightarrow & (\forall r_i, r_j. r_i \leq r_j \in \Omega \rightarrow M(r_i) \subseteq M(r_j))\ \wedge \\
              &                 & (\forall r_i, r_j. r_i * r_j \in \Omega \rightarrow M(r_i) \cap M(r_j) = \emptyset) \\
L \sim_H \Gamma & \Leftrightarrow & \forall (id,v) \in L. v \sim_H M \llbracket \Gamma \rrbracket (id) \\
S \sim H & \Leftrightarrow & \forall (l,v) \in S. v \sim_H H(l) \\
\end{array} & 
\begin{array}{lll}
E \sim H & \Leftrightarrow & (\forall l, c, v. write(l,c, v) \in E \rightarrow v \sim_H H(l))\ \wedge \\
& & (\forall id, l, v. reduce_{id}(l, v) \in E \rightarrow \\
& & (M \llbracket \Gamma \rrbracket (id) = (\hat T_1, \hat T_2), \emptyset, \emptyset \rightarrow \hat T_1) \wedge H(l) = \hat T_1 \wedge v \sim_H \hat T_2) \\
\end{array}
\end{array} \vspace{3mm} \\
\begin{array}{l@{\hspace{1cm}}l}
\begin{array}{l}
bv \sim_H bool \\
iv \sim_H int \\
null \sim_H \hat T@\rho \\
\end{array} &
\begin{array}{lll}
l \sim_H \hat T@\rho & \Leftrightarrow & l \in \rho \wedge H(l) \sim_H \hat T \\
\langle v_1, v_2 \rangle \sim_H \langle \hat T_1, \hat T_2 \rangle & \Leftrightarrow & (v_1 \sim_H \hat T_1) \wedge (v_2 \sim_H \hat T_2) \\
\langle \langle \oton{\rho}, v \rangle \rangle \sim_H \text{rr}[\oton{r}] \hat T\text{ where }\Omega & \Leftrightarrow & (v \sim_H \hat T[\rho_1/r_1, \ldots \rho_n/r_n]) \wedge (\{ (r_i, \rho_i) \} \sim \Omega) \\
K \sim_H coloring(\rho) & \Leftrightarrow & \forall l_1, v_1. (l_1, v_1) \in K \rightarrow (l_1 \in \rho\ \wedge \\
& & \forall l_2, v_2. (l_2, v_2) \in K \rightarrow (l_1 \not= l_2) \vee (v_1 = v_2)) \\
\end{array}
\end{array}
\end{array}
$
}
\caption{Consistency Properties}
\label{fig:constprop}
\end{figure*}

\begin{thm}
For any expression $e$, if:
\begin{center}
$
\begin{array}{l} \\
\typeenv{e}{T} \\
\opsenv{e}{v}{E} \\
M \sim \Omega \\
L \sim_H \Gamma \\
S \sim H
\end{array}
$
\end{center}
then:
\begin{center}
$
\begin{array}{l}
v \sim_H M \llbracket T \rrbracket \\
E \sim H \\
M \vdash E : \Phi
\end{array}
$
\end{center}
\end{thm}

Although there are 23 different cases to be proven, many of them are similar, and benefit from
the use of the following lemmas:

\begin{lem}
\label{lemma:heapconst:apply}
If $S \sim H$ and $E \sim H$, then $\text{apply}(S, E, C) \sim H$.
\end{lem}

\begin{lem}
\label{lemma:heapconst:effects1}
If $E_1 \sim H$ and $E_2 \sim H$, then $E_1 \dplus E_2 \sim H$.
\end{lem}

\begin{lem}
\label{lemma:heapconst:effects2}
If $E_1 \sim H$ and $E_2 \sim H$ and $valid\_interleave(S, C, E', E_1, E_2)$, then $E' \sim H$.
\end{lem}

\begin{lem}
\label{lemma:effsound:effects1}
If $M \vdash E_1 : \Phi$ and $M \vdash E_2 : \Phi$, then $M \vdash E_1 \dplus E_2 : \Phi$.
\end{lem}

\begin{lem}
\label{lemma:effsound:effects2}
If $M \vdash E_1 : \Phi$ and $M \vdash E_2 : \Phi$ and $valid\_interleave(S, C, E', E_1, E_2)$, then $M \vdash E' : \Phi$.
\end{lem}

\begin{lem}
\label{lemma:omegaclosure}
$M \sim \Omega^*$ if and only if $M \sim \Omega$.
\end{lem}

\begin{lem}
\label{lemma:phiclosure}
$M \vdash E : \Phi^*$ if ant only if $M \vdash E : \Phi$.
\end{lem}

The complete soundness proof can be found here\cite{xyz} - for space reasons we just discuss
the general strategy here.

\begin{itemize}

\item Only three expressions have subexpressions that modify $M$ or $\Omega$.  For the
{\em partition} expression, the assumed consistency of the coloring can be shown to preserve
the region mapping consistency with respect to the constraints.  Similarly, the consistency of 
a region relation instance guarantees the maintenance of consistency in an {\em unpack}
expression.  Finally, the body of a called function executes with a different region mapping,
but with an initially-empty set of constraints, so they are trivially satisfied.

\item Only three expression have subexpressions that modify $L$, $\Gamma$, $M$.  The
{\em partition} expression only modifies $M$, and the requirement that it not reuse existing
names ensures that $M \llbracket \Gamma \rrbracket$ does not change.  The value and type of the
binding created in a {\em let} expression is obviously consistent, while the binding created in
an {\em unpack} expression is less obviously so, requiring an induction over the type of the 
unpacked value to show equivalence under the new mapping.  The last case is again the body of a
called function, which requires the same style of proof as for the {\em unpack} expression to
be performed for each formal parameter.

\item The heap typing consistency of all stores used in subexpressions follows directly from
Lemma~\ref{lemma:heapconst:apply}.

\item With the necessary consistency conditions all satified, the inductive hypothesis asserts
the type consistency of all subexpression result values.  The consistency of the result of 
the {\em upregion} is guaranteed by the type checking requirement of appropriate subregion
constraints and the mapping's consistency with those constraints, whereas {\em downregion}
expression's result is consistent due to the run time check that is performed.  (The requirement
for appropriate constraints in the {\em downregion} typing rule is primarily for performance
(i.e. knowing which part of the region tree to examine) and to catch downcasts that will never
work at runtime.)

The consistency of a {\em read} expression's result is trivial when an address falls in the
clobber set, and counts on the consistency of the store otherwise.  The consistency of a 
{\em color} expression's result depends on the pointer subexpression's consistency and the 
removal of any previous coloring of that location from the coloring set.  The remaining
interesting cases arise from changes to the mapping $M$ rather than transformations on the
value $v$.  In the case of {\em partition} and {\em unpack} expressions, the type system's
requirement that the body subexpression's result not use the regions that were added to the
mapping allow the changes to the mapping to be ignored.  The last case is, yet again, the body
of a called function, and the same strategy that was used for the type consistency of the
formal parameters works in reverse for the function's result.

\item The type consistency of the value in an expression's effect follows from
Lemmas~\ref{lemma:heapconst:effects1} and~\ref{lemma:heapconst:effects2}.  New effects are
added by the {\em write} and {\em reduce} expressions, but their values' consistency follows
directly from the inductive hypothesis on their subexpressions.  Finally, the consistency of the
values in a called function's memory trace is addressed in the same way as the return value.

\item The proof of the crucial property of containment of effects within the available privileges
is very similar to the previous step, with Lemmas~\ref{lemma:effsound:effects1}
and~\ref{lemma:effsound:effects2} covering most cases, straightforward proofs for the {\em read},
{\em write}, and {\em reduce} expressions, and one final special case for function calls.  The 
inductive hypothesis asserts that the function body's effects are contained within its own
privileges.  Those are a subset of the caller's $\Phi^*$, and then Lemma~\ref{lemma:phiclosure}
is used to complete the proof.
\end{itemize}

modify $M$
\begin{table*}
\centering
{\small
\begin{tabular}{l|p{5cm}|p{5cm}|p{5cm}}
{\bf Rule} & 
\multicolumn{1}{|c|}{$\mathbf{v \sim_H M \llbracket T \rrbracket}$} &
\multicolumn{1}{|c|}{$\mathbf{E \sim H}$} &
\multicolumn{1}{|c|}{$\mathbf{E : \Phi}$} \\
\hline
BoolConst & & & \\
IntConst & & & \\
TupleBuild & & & \\
TupleField & & & \\
LocalVar & & & \\
NullPtr & & & \\
NewPtr & & & \\
IsNull & & & \\
UpRegion & & & \\
DownRegion & & & \\
Read & & & \\
Write & & & \\
Reduce & & & \\
Newcolor & & & \\
Color & & & \\
IntMath & & & \\
IntCmp & & & \\
Let & & & \\
IfElse & & & \\
TaskCall & & & \\
Partition & & & \\
Pack & & & \\
Unpack & & &
\end{tabular}}
\caption{Proof pieces}
\end{table*}
